---
title: "The Story Behind Game Playtimes"
---

## About HowLongToBeat

**[HowLongToBeat](https://howlongtobeat.com)** is a community-driven database that crowdsources video game completion times from millions of players worldwide. Instead of relying on a single playtime estimate, the platform breaks play styles down into three core categories:

- **Main Story**: Finishing the core objectives to see the end credits without straying into side content.
- **Main + Extras**: Completing the main storyline while taking time to explore side quests, optional bosses, and character extras.
- **100% Completionist**: Doing everything possible in the game—unlocking every achievement/trophy, discovering all secrets, and collecting 100% of items.

Our database contains playtime and rating metrics for over **3,100 of the most popular video games**, ranging from retro classics to modern releases.

---

## Have Games Really Gotten Longer?

A frequent debate among gamers is whether modern titles demand far more time than older generations. By tracking median completion times across release years (2000–2024), we can see how game lengths have evolved across all three play styles.

*Hover over any year on the chart below to inspect playtimes simultaneously, or click a style in the legend to toggle it:*

```{r}
#| message: false
#| echo: false

library(tidyverse)
library(plotly)

# Load data and compute annual medians for each playstyle
games <- read_csv("howlongtobeat.csv", show_col_types = FALSE)

yearly <- games |>
  filter(!is.na(release_year), !is.na(main_story_hours), !is.na(completionist_hours), release_year >= 2000, release_year <= 2024) |>
  summarize(
    `Main Story` = round(median(main_story_hours, na.rm = TRUE), 1),
    `Main + Extras` = round(median(main_extra_hours, na.rm = TRUE), 1),
    `100% Completionist` = round(median(completionist_hours, na.rm = TRUE), 1),
    .by = release_year
  ) |>
  arrange(release_year) |>
  pivot_longer(
    cols = c(`Main Story`, `Main + Extras`, `100% Completionist`),
    names_to = "Playstyle",
    values_to = "Hours"
  )

# Create line chart
p <- yearly |>
  ggplot(aes(x = release_year, y = Hours, color = Playstyle, group = Playstyle)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c(
    "Main Story" = "#3498db",
    "Main + Extras" = "#f39c12",
    "100% Completionist" = "#e74c3c"
  )) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 4)) +
  labs(
    title = "Median Hours to Beat by Play Style (2000–2024)",
    x = "Release Year",
    y = "Median Playtime (Hours)"
  ) +
  theme_minimal() +
  theme(
    plot.background = element_rect(fill = "#222222", color = NA),
    panel.background = element_rect(fill = "#222222", color = NA),
    text = element_text(color = "#ffffff"),
    axis.text = element_text(color = "#ffffff"),
    legend.background = element_rect(fill = "#222222", color = NA)
  )

ggplotly(p) |>
  layout(
    paper_bgcolor = "#222222",
    plot_bgcolor = "#222222",
    font = list(color = "#ffffff"),
    legend = list(font = list(color = "#ffffff")),
    hovermode = "x unified"
  )
```

---

### Key Takeaways

1. **Main Stories Are Remarkably Stable**: For over two decades, the core campaign length of popular games has consistently hovered around **11 to 16 hours**.
2. **Completionist Time Has Surged**: While the main campaign hasn't ballooned, the depth of side content and 100% completionist runs has grown significantly with modern open-world game design.

---

## The Backlog Trap: Bought vs. Finished

Every gamer knows the feeling of buying an acclaimed title on sale, only for it to sit untouched in their backlog.

The chart below shows the **percentage of players who finished the game** (blue) versus those who **still have it in their backlog** (red) for the most backlogged titles in gaming history.

- **Blue (Finished)**: Games players actually see through to the credits (like *Portal 2* at over 60%).
- **Red (Backlogged)**: Games people bought with great intentions, but left unfinished (like *BioShock 2* and *Fallout: New Vegas* where over 60% remain uncompleted).
- **Dashed line**: The 50/50 mark.

*Hover over any bar to see exact percentages and player counts:*

```{r}
#| message: false
#| echo: false

# Select games with large community counts (20k+) and compute finish vs backlog shares
top_backlogged <- games |>
  filter(!is.na(count_backlog), !is.na(count_comp), (count_backlog + count_comp) >= 20000) |>
  mutate(
    total = count_backlog + count_comp,
    `Finished (%)` = round((count_comp / total) * 100, 1),
    `Unfinished / Backlog (%)` = round((count_backlog / total) * 100, 1)
  ) |>
  arrange(desc(count_backlog)) |>
  slice_head(n = 12) |>
  select(game_name, `Finished (%)`, `Unfinished / Backlog (%)`) |>
  pivot_longer(
    cols = c(`Finished (%)`, `Unfinished / Backlog (%)`),
    names_to = "Status",
    values_to = "Percent"
  )

p_backlog <- top_backlogged |>
  ggplot(aes(x = reorder(game_name, Percent), y = Percent, fill = Status)) +
  geom_col(position = "fill") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c(
    "Finished (%)" = "#3498db",
    "Unfinished / Backlog (%)" = "#e74c3c"
  )) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "#ffffff") +
  labs(
    title = "Top Backlogged Games: Finished vs. Left in Backlog",
    x = NULL,
    y = "Share of Players"
  ) +
  theme_minimal() +
  theme(
    plot.background = element_rect(fill = "#222222", color = NA),
    panel.background = element_rect(fill = "#222222", color = NA),
    text = element_text(color = "#ffffff"),
    axis.text = element_text(color = "#ffffff"),
    legend.background = element_rect(fill = "#222222", color = NA)
  )

ggplotly(p_backlog) |>
  layout(
    paper_bgcolor = "#222222",
    plot_bgcolor = "#222222",
    font = list(color = "#ffffff"),
    legend = list(font = list(color = "#ffffff"), orientation = "h", y = -0.15)
  )
```
