---
title: "The Story Behind Game Playtimes"
---

## About HowLongToBeat

**[HowLongToBeat](https://howlongtobeat.com)** is a community-driven database that crowdsources video game completion times from millions of players worldwide. Instead of relying on a single playtime estimate, the platform breaks play styles down into three core categories:

- **Main Story**: Finishing the core objectives to see the end credits without straying into side content.
- **Main + Extras**: Completing the main storyline while taking time to explore side quests, optional bosses, and character extras.
- **100% Completionist**: Doing everything possible in the game—unlocking every achievement/trophy, discovering all secrets, and collecting 100% of items.

Our database contains playtime and rating metrics for over **3,100 of the most popular video games**, ranging from retro classics to modern releases.

---

## Have Games Really Gotten Longer?

A frequent debate among gamers is whether modern titles demand far more time than older generations. By tracking median completion times across release years (2000–2024), we can see how game lengths have evolved across all three play styles.

*Hover over any year on the chart below to inspect playtimes simultaneously, or click a style in the legend to toggle it:*

```{r}
#| message: false
#| echo: false

library(tidyverse)
library(plotly)

# Load data and compute annual medians for each playstyle
games <- read_csv("howlongtobeat.csv", show_col_types = FALSE)

yearly <- games |>
  filter(!is.na(release_year), !is.na(main_story_hours), !is.na(completionist_hours), release_year >= 2000, release_year <= 2024) |>
  summarize(
    `Main Story` = round(median(main_story_hours, na.rm = TRUE), 1),
    `Main + Extras` = round(median(main_extra_hours, na.rm = TRUE), 1),
    `100% Completionist` = round(median(completionist_hours, na.rm = TRUE), 1),
    .by = release_year
  ) |>
  arrange(release_year) |>
  pivot_longer(
    cols = c(`Main Story`, `Main + Extras`, `100% Completionist`),
    names_to = "Playstyle",
    values_to = "Hours"
  )

# Create line chart
p <- yearly |>
  ggplot(aes(x = release_year, y = Hours, color = Playstyle, group = Playstyle)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c(
    "Main Story" = "#3498db",
    "Main + Extras" = "#f39c12",
    "100% Completionist" = "#e74c3c"
  )) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 4)) +
  labs(
    title = "Median Hours to Beat by Play Style (2000–2024)",
    x = "Release Year",
    y = "Median Playtime (Hours)"
  ) +
  theme_minimal() +
  theme(
    plot.background = element_rect(fill = "#222222", color = NA),
    panel.background = element_rect(fill = "#222222", color = NA),
    text = element_text(color = "#ffffff"),
    axis.text = element_text(color = "#ffffff"),
    legend.background = element_rect(fill = "#222222", color = NA)
  )

ggplotly(p) |>
  layout(
    paper_bgcolor = "#222222",
    plot_bgcolor = "#222222",
    font = list(color = "#ffffff"),
    legend = list(font = list(color = "#ffffff")),
    hovermode = "x unified"
  )
```

---

### Key Takeaways

1. **Main Stories Are Remarkably Stable**: For over two decades, the core campaign length of popular games has consistently hovered around **11 to 16 hours**.
2. **Completionist Time Has Surged**: While the main campaign hasn't ballooned, the depth of side content and 100% completionist runs has grown significantly with modern open-world game design.

---

## Finished vs. Unfinished Games

Every gamer knows the feeling of starting an acclaimed game, only to leave it unfinished.

The chart below shows the **percentage of players who finished the game** (blue) versus those who left it **unfinished** (red) for some of the most widely owned titles.

- **Blue (Finished)**: Players who saw the game through to the end credits.
- **Red (Unfinished)**: Players who left the game uncompleted.
- **Dashed line**: The 50/50 mark.

*Hover over any bar to see exact percentages:*

```{r}
#| message: false
#| echo: false

# Select games with large community counts (20k+) and compute finished vs unfinished shares
top_backlogged <- games |>
  filter(!is.na(count_backlog), !is.na(count_comp), (count_backlog + count_comp) >= 20000) |>
  mutate(
    total = count_backlog + count_comp,
    `Finished (%)` = round((count_comp / total) * 100, 1),
    `Unfinished (%)` = round((count_backlog / total) * 100, 1)
  ) |>
  arrange(desc(count_backlog)) |>
  slice_head(n = 12) |>
  select(game_name, `Finished (%)`, `Unfinished (%)`) |>
  pivot_longer(
    cols = c(`Finished (%)`, `Unfinished (%)`),
    names_to = "Status",
    values_to = "Percent"
  )

p_backlog <- top_backlogged |>
  ggplot(aes(x = reorder(game_name, Percent), y = Percent, fill = Status)) +
  geom_col(position = "fill") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c(
    "Finished (%)" = "#3498db",
    "Unfinished (%)" = "#e74c3c"
  )) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "#ffffff") +
  labs(
    title = "Popular Games: Finished vs. Unfinished",
    x = NULL,
    y = "Share of Players"
  ) +
  theme_minimal() +
  theme(
    plot.background = element_rect(fill = "#222222", color = NA),
    panel.background = element_rect(fill = "#222222", color = NA),
    text = element_text(color = "#ffffff"),
    axis.text = element_text(color = "#ffffff"),
    legend.background = element_rect(fill = "#222222", color = NA)
  )

ggplotly(p_backlog) |>
  layout(
    paper_bgcolor = "#222222",
    plot_bgcolor = "#222222",
    font = list(color = "#ffffff"),
    legend = list(font = list(color = "#ffffff"), orientation = "h", y = -0.15)
  )
```

---

## Main Story vs. 100% Completionist (Top Games)

```{r}
#| message: false
#| echo: false

# Select top popular games and reshape for comparison
top_games <- games |>
  filter(!is.na(main_story_hours), !is.na(completionist_hours)) |>
  slice_head(n = 12) |>
  select(game_name, `Main Story` = main_story_hours, `Completionist` = completionist_hours) |>
  pivot_longer(cols = c(`Main Story`, `Completionist`), names_to = "Playstyle", values_to = "Hours")

p_bars <- top_games |>
  ggplot(aes(x = reorder(game_name, Hours), y = Hours, fill = Playstyle)) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = c("Main Story" = "#375a7f", "Completionist" = "#e74c3c")) +
  labs(
    title = "Main Story vs. 100% Completionist (Top Games)",
    x = NULL,
    y = "Hours to Beat"
  ) +
  theme_minimal() +
  theme(
    plot.background = element_rect(fill = "#222222", color = NA),
    panel.background = element_rect(fill = "#222222", color = NA),
    text = element_text(color = "#ffffff"),
    axis.text = element_text(color = "#ffffff"),
    legend.background = element_rect(fill = "#222222", color = NA)
  )

ggplotly(p_bars) |>
  layout(
    paper_bgcolor = "#222222",
    plot_bgcolor = "#222222",
    font = list(color = "#ffffff"),
    legend = list(font = list(color = "#ffffff"))
  )
```

---

## Where Does Your Time Go? (2000–2026)

When aiming for 100% completion, how is player time divided between the main campaign, side content, and completionist grinding?

```{r}
#| message: false
#| echo: false

# Calculate average time breakdown across 2000-2026 titles
breakdown <- games |>
  filter(
    !is.na(release_year),
    release_year >= 2000,
    release_year <= 2026,
    !is.na(main_story_hours),
    !is.na(main_extra_hours),
    !is.na(completionist_hours),
    main_extra_hours >= main_story_hours,
    completionist_hours >= main_extra_hours
  ) |>
  summarize(
    `Main Story` = round(mean(main_story_hours), 1),
    `Side Quests & Extras` = round(mean(main_extra_hours - main_story_hours), 1),
    `100% Collectibles & Secrets` = round(mean(completionist_hours - main_extra_hours), 1)
  ) |>
  pivot_longer(everything(), names_to = "Category", values_to = "Hours")

# Animated interactive donut chart
plot_ly(
  breakdown,
  labels = ~Category,
  values = ~Hours,
  type = "pie",
  hole = 0.58,
  textinfo = "label+percent",
  textposition = "outside",
  marker = list(
    colors = c("#3498db", "#f39c12", "#e74c3c"),
    line = list(color = "#222222", width = 2)
  ),
  hoverinfo = "text",
  hovertext = ~paste0("<b>", Category, "</b><br>Average: ", Hours, " hrs")
) |>
  layout(
    title = list(text = "Average Playtime Distribution (2000–2026)", font = list(color = "#ffffff")),
    paper_bgcolor = "#222222",
    plot_bgcolor = "#222222",
    font = list(color = "#ffffff"),
    showlegend = TRUE,
    legend = list(orientation = "h", x = 0.05, y = -0.15, font = list(color = "#ffffff"))
  )
```

